---
title: "Project 1"
authors: Mary Gibbs and Rayhaan Rasheed
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

Research Question: Do certain genres of video games sell better in different regions (North America, Europe, and Japan)?

H0: Each genre of video game does sell the same across different regions (North America, Europe, Japan).

HA: Each genre of video game does not sell the same across different regions (North America, Europe, Japan).

Packages
```{r}
library(data.table)
library(dplyr)
library(ggplot2)
library(plyr)
```

Read in data
```{r}
video_games <- fread("video_games.csv")
```

Check data
```{r}
str(video_games)
```

Clean data
```{r}
#create data frame with columns of interest
vg <- video_games[, c(1, 4, 6:8, 10, 11, 13)]
#correctly classify variables
vg$Critic_Score <- as.numeric(vg$Critic_Score)
vg$User_Score <- as.numeric(vg$User_Score)
#remove NA rows
vg <- na.omit(vg)
#multiply User_Score by 10 to match Critic_Score 
vg$User_Score <- (vg$User_Score)*10
#check data 
str(vg)
```

EDA
```{r}
#vg summary
summary(vg)
```

```{r}
#
```

