---
title: "Project 1"
authors: Mary Gibbs and Rayhaan Rasheed
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

Research Question: Do certain genres of video games sell better in different regions (North America, Europe, and Japan)?

H0: Each genre of video game does sell the same across different regions (North America, Europe, Japan).

HA: Each genre of video game does not sell the same across different regions (North America, Europe, Japan).

Packages
```{r}
library(data.table)
library(dplyr)
library(ggplot2)
library(plyr)
```

Read in data
```{r}
video_games <- fread("video_games.csv")
```

Check data
```{r}
str(video_games)
```

Clean data
```{r}
#create data frame with columns of interest
vg <- video_games[, c(1, 4, 6:8, 10, 11, 13)]
#correctly classify variables
vg$Critic_Score <- as.numeric(vg$Critic_Score)
vg$User_Score <- as.numeric(vg$User_Score)
#remove NA rows
vg <- na.omit(vg)
#multiply User_Score by 10 to match Critic_Score 
vg$User_Score <- (vg$User_Score)*10
#check data 
str(vg)
```

EDA
```{r}
#vg summary
summary(vg)
```
```{r}
#boxplot NA_Sales
boxplot(vg$NA_Sales)
#remove major outliers
NA_Sales_no_mo <- vg$NA_Sales[!vg$NA_Sales %in% boxplot.stats(vg$NA_Sales)$out]
#boxplot NA_Sales_no_mo
boxplot(NA_Sales_no_mo)
```
```{r}
#histogram NA_Sales_no_mo
hist(NA_Sales_no_mo, breaks = 10)
```

```{r}
#boxplot EU_Sales
boxplot(vg$EU_Sales)
#remove major outliers
EU_Sales_no_mo <- vg$EU_Sales[!vg$EU_Sales %in% boxplot.stats(vg$EU_Sales)$out]
#boxplot EU_Sales_no_mo
boxplot(EU_Sales_no_mo)
```

```{r}
#histogram EU_Sales_no_mo
hist(EU_Sales_no_mo, breaks = 10)
```

```{r}
#boxplot JP_Sales
boxplot(vg$JP_Sales)
#remove major outliers
JP_Sales_no_mo <- vg$JP_Sales[!vg$JP_Sales %in% boxplot.stats(vg$JP_Sales)$out]
#boxplot NA_Sales_no_mo
boxplot(JP_Sales_no_mo)
```

```{r}
#histogram JP_Sales_no_mo
hist(JP_Sales_no_mo, breaks = 10)
```

```{r}
#split points and outliers into two different data frames

```

