---
title: "Analysis of Education Expenditure and 8th Grade Average Math Scores in the United States from 2005-2015"
course: DATS 6101 Introduction to Data Science
authors: Mary Gibbs and Rayhaan Rasheed
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

----------------
Project Overview
----------------

Research Question: Is there a statistically significant relationship between education expenditure and 8th grade average math scores in the U.S. from 2005 to 2015?

Hypotheses:
H0: There is no statistically significant relationship between education expenditure and 8th grade average math scores in the U.S. from 2005 to 2015. β1 = 0. 
HA: There is a statistically significant relationship between education expenditure and 8th grade average math scores in the U.S. from 2005 to 2015. β1 ≠ 0.

Data Source: https://www.kaggle.com/noriuk/us-educational-finances

----------------
Data Preparation
----------------

Packages
```{r}
library(corrplot)
library(data.table)
library(dplyr)
library(ggfortify)
library(ggplot2)
library(ggvis)
library(grid)
library(plyr)
library(psych)
library(stats)
library(VIM)
```

Read in data
```{r}
finances <- fread("finances.csv")
scores <- fread("scores.csv")
```

Check data
```{r}
str(finances)
str(scores)
```

Clean data
```{r}
#choose columns of interest
finances <- finances[, c(1, 2, 4, 8)]
#join finances and scores on year and state
df <- merge(finances, scores, by = c("YEAR", "STATE"))
#make column names lowercase
df <- setnames(df, tolower(names(df)))
#select 2005-2015
df <- df[df$year > 2004 & df$year < 2016]
#get total revenue in USD (millions)) and total expenditure in USD (millions)
df$total_revenue <- df$total_revenue/1000000
df$total_expenditure <- df$total_expenditure/1000000
#remove rows with avg_score = "-", avg_score = "‡"
df <- df[df$avg_score != "—" & df$avg_score != "‡"]
#select mathematics
df <- df[df$test_subject == 'Mathematics']
#select 8th grade
df <- df[df$test_year == 8]
#change year, state, avg_score, and test_year data types
df$year <- as.factor(df$year)
df$state <- as.factor(df$state)
df$avg_score <- as.numeric(df$avg_score, digits = 7)
df$test_year <- as.character(df$test_year)
#check data frame 
str(df)
```

-------------------------
Exploratory Data Analysis
-------------------------

Look at total revenue, total expenditure, and 8th grade average math scores from 2005-2015
```{r}
#get total revenue sum, total expenditure sum, and 8th grade average math scores sum from 2005-2015
df_tr_te_sum <- aggregate(list(df$total_revenue, df$total_expenditure), by = list(df$year), FUN = sum)
colnames(df_tr_te_sum) <- c("year", "total_revenue_sum", "total_expenditure_sum")
df_tr_te_sum <- melt(df_tr_te_sum, id.vars = "year")
df_avg_score_sum <- aggregate(list(df$avg_score), by = list(df$year), FUN = sum)
colnames(df_avg_score_sum) <- c("year", "avg_score_sum")
df_avg_score_sum <- melt(df_avg_score_sum, id.vars = "year")
#line plot of total revenue sum, total expenditure sum, and 8th grade average math scores sum from 2005-2015
ggplot(df_tr_te_sum, aes(x = year, y = value, group = variable, colour = variable)) +
      geom_line() + 
      ggtitle ("Total Revenue Sum and Total Expenditure Sum") +
      theme(plot.title = element_text(hjust = 0.5)) +
      xlab("Year") +
      ylab("USD (millions)") + 
      labs(color = "Variable") + 
      scale_color_manual(values = c("darkolivegreen3", "deepskyblue2"))
ggplot(df_avg_score_sum, aes(x = year, y = value, group = variable, colour = variable)) +
      geom_line() + 
      ggtitle ("8th Grade Average Math Scores Sum") +
      theme(plot.title = element_text(hjust = 0.5)) + 
      xlab("Year") +
      ylab("Average Scores Sum") + 
      labs(color = "Variable") + 
      scale_color_manual(values = c("darkorchid3"))
```

Look at sum of total revenue - total expenditure from 2005-2015
```{r}
#get sum of total revenue - total expenditure from 2005-2015
df$tr_minus_te <- df$total_revenue - df$total_expenditure
df_sum_tr_minus_te <- aggregate(list(df$tr_minus_te), by = list(df$year), FUN = sum)
colnames(df_sum_tr_minus_te) <- c("year", "sum_total_revenue-total_expenditure")
df_sum_tr_minus_te <- melt(df_sum_tr_minus_te, id.vars = "year")
#line plot of sum of total revenue - total expenditure from 2005-2015
ggplot(df_sum_tr_minus_te, aes(x = year, y = value, group = variable, colour = variable)) +
      geom_line() + 
      ggtitle ("Sum of Total Revenue - Total Expenditure") +
      theme(plot.title = element_text(hjust = 0.5)) +
      xlab("Year") +
      ylab("USD (millions)") + 
      labs(color = "Variable") + 
      scale_color_manual(values = c("darkorange2"))
```

Descriptive statistics
```{r}
#describe data frame
describe(df[, 3:5])
```
```{r}
#summarize data frame
summary(df)
```

Look at outliers
```{r}
#divide graph area into three columns
par(mfrow = c(1, 3))  
#boxplots of total revenue, total expenditure, and 8th grade average math scores
boxplot(df$total_revenue, main = "Total Revenue", ylab = "USD (millions)")
boxplot(df$total_expenditure, main = "Total Expenditure", ylab = "USD (millions)")  
boxplot(df$avg_score, main = "8th Grade Average Math Scores", ylab = "Average Score")
#total revenue and total expenditure have outliers

#remove total revenue and total expenditure outliers 
tr_outliers <- boxplot(df$total_revenue, plot = FALSE)$out
te_outliers <- boxplot(df$total_expenditure, plot = FALSE)$out
df <- df[!df$total_expenditure %in% c(tr_outliers, te_outliers)]
#divide graph area into three columns
par(mfrow = c(1, 2))  
#boxplots of total revenue (no outliers) and total expenditure (no outliers)
boxplot(df$total_revenue, main = "Total Revenue (No Outliers)", ylab = "USD (millions)")
boxplot(df$total_expenditure, main = "Total Expenditure (No Outliers)", ylab = "USD (millions)")
#total revenue and total expenditure have no outliers
```

Look at normality 
```{r}
#divide graph area into three columns
par(mfrow = c(1, 3))  
#density plots of total revenue, total expenditure, and 8th grade average math scores
plot(density(df$total_revenue), main = "Total Revenue", xlab = "USD (millions)", ylab = "Frequency", col = "darkolivegreen3")
polygon(density(df$total_revenue), col = "darkolivegreen3")
plot(density(df$total_expenditure), main = "Total Expenditure", xlab = "USD (millions)", ylab = "Frequency", col = "deepskyblue2")
polygon(density(df$total_expenditure), col = "deepskyblue2")
plot(density(df$avg_score), main = "8th Grade Average Math Scores", xlab = "Average Score", ylab = "Frequency", col = "darkorchid3")
polygon(density(df$avg_score), col = "darkorchid3")
#total revenue and total expenditure have slight left skew, 8th grade average math scores looks normally distributed 

#Shapiro-Wilk normality test
shapiro.test(df$total_revenue)
shapiro.test(df$total_expenditure)
shapiro.test(df$avg_score)
#total revenue, total expenditure, and 8th grade average math scores have p values < 0.05, not normally distributed 
```

Look at linearity 
```{r}
#scatter plot 
ggplot(df, aes(x = total_expenditure, y = avg_score)) +
      geom_point(shape = 1) +    
      geom_smooth(method = lm, se = FALSE, color = "goldenrod1") + 
      ggtitle("8th Grade Average Math Test Scores vs. Total Expenditure") + 
      theme(plot.title = element_text(hjust = 0.5)) + 
      xlab("Total Expenditure (USD (millions))") + 
      ylab("Average Score")
#linearity 
```

Look at correlations
```{r}
#covariance matrix 
df_cov <- cov(df[, 3:5], method = "pearson")
df_cov
#correlation matrix 
df_cor <- cor(df[, 3:5], method = "pearson")
df_cor
#correlation plot 
corrplot(df_cor, method = "square", addCoef.col = "black", number.digits = 4)
#all relationships are positive
#strong positive relationship between total revenue and total expenditure
#weak positive relationships between total revenue and 8th grade average math scores and total expenditure and 8th grade average math scores
```

-----
Model
-----

Linear regression of 8th grade average math scores vs. total expenditure
```{r}
#model
df_lm <- lm(avg_score ~ total_expenditure, data = df)
summary(df_lm)
#residuals - not normally distributed
#coefficients - total_expenditure p-value < 0.05, statistically significant relationship between school expenditure and 8th grade average math scores
#adjusted R-squared - 3.59% of the variance in 8th grade average math scores can be explained by school expenditure, poor model

#model assumptions 
par(mfrow = c(2, 2))
plot(df_lm)
df_lm_res <- df_lm$residuals
describe(df_lm_res)
shapiro.test(df_lm_res)
#linearity - linearity, scatterplot shows linearity, residual vs fitted plot shows residuals skewed to left and spread around nearly horizontal line
#independence of residuals - independence of residuals, observations are independent
#normality of residuals - no normality of residuals, normal Q-Q plot shows that residuals follow straight line well, Shapiro-Wilk normality test shows p value < 0.05
#homoscedasticity of residuals - possible homoscedasticity of residuals, residual vs fitted plot shows residuals skewed to left and spread around slightly negatively sloped line
```
